AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction

### 2017-09-08 doc/ms.tex
Elsevier format
Version downloaded from Overleaf

\documentclass[a4paper, 12pt]{article}
\usepackage{amsmath}
\usepackage{color}
\usepackage{dsfont}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{pdflscape}

@@

\title{Crystals}
\author{Brendon J. Brewer}
\date{}

\begin{document}
\maketitle

%\abstract{\noindent Abstract}

% Need this after the abstract
\setlength{\parindent}{0pt}
\setlength{\parskip}{8pt}

\section{Bayesian inference}
In the Bayesian framework
\citep{gregory2005bayesian, o2004kendall, sivia2006data},
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Typically, one starts with the ``prior distribution''
for unknown quantities (`parameters') $\params$, written $p(\params | \info)$
(the probability distribution for $\params$ {\em given} prior information
$\info$). This is supplemented with $p(\data | \params, \info)$,
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. This describes uncertainty about what data will be observed,
but knowledge about the fact that the data has something to do with the
parameters $\params$.
The
product of these two distributions is the joint prior
\begin{align}
p(\params, \data | \info) &= p(\params | \info)p(\data | \params, \info)
\end{align}
which describes uncertainty about the value of both the parameters and the
dataset.

Once a specific observed dataset $\data_o$ is known, knowledge about
$\params$ is updated from the prior to the posterior
\begin{align}
p(\params | \data_0, \info) &\propto
    p(\params | \info)p(\data_o | \params, \info).
\end{align}
which takes the specific data $\data_0$ into account. With a specific dataset
$\data_0$ being specified, $p(\data_o | \params,\info)$ is a function of the
unknown parameters $\params$ and is called the {\em likelihood function}.
The posterior distribution is therefore proportional to the prior distribution
times the likelihood function.

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\theta$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\theta > 3.4$.

We compute the posterior distribution for the unknown parameters of the
components, using \citep{dnest4} to generate samples from that posterior
distribution.

\section{The model curve}
The algorithm fits the data with a model curve
$f(\x)$, defined between $\x=\x_{\rm min}$ and $\x=\x_{\rm max}$,
which is assumed to be a sum of the following components:
\begin{enumerate}
\item A background component, described by parameters $\params_{\rm bg}$
\item A wide component, described by parameters $\params_{\rm wide}$
\item Narrow spikes, described by parameters $\params_{\rm narrow}$
\end{enumerate}

The total model function $f_{\rm tot}(x)$ is therefore
\begin{align}
f_{\rm tot}(\x) &= f_{\rm bg}(\x; \params_{\rm bg})
       + f_{\rm wide}(\x; \params_{\rm wide})
       + f_{\rm narrow}(\x; \params_{\rm narrow})
\end{align}
and the crystallinity is defined as the proportion of the area in the
wide and narrow components which is accounted for by the wide component:
\begin{align}
C &= \frac{\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm wide}(\x; \params_{\rm wide}) \, d\x}
          {\int_{\x_{\rm min}}^{\x_{\rm max}}
            \left[f_{\rm wide}(\x; \params_{\rm wide})
            + f_{\rm narrow}(\x; \params_{\rm narrow}) \right] \, d\x}
\end{align}

From each sample,
it is possible to compute the crystallinity, making it straightforward to
calculate the marginal posterior distribution of the crystallinity, describing
the remaining uncertainty about its value in the light of the data.

In the following subsections, I describe the specific parameterisations of the
three parts of the model curve.

\subsection{The background component}
The background component is assumed to be piecewise linear
{\todo between $\x=\x_{\rm min}$ and $\x=10$, between $\x=10$ and $\x=40$,
and between $\x=40$ and $\x=\x_{\rm max}$}.
We parameterise the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which set the positions of four control points:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right) \nonumber
\end{align}
The value of the background component at any other point is found by
linear interpolation. Figure~\ref{fig:background} shows three example
background curves generated from this parameterisation, with typical
values of the parameters.

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figures/background.pdf}
\caption{\it Three example functions showing possible shapes of the
background component.\label{fig:background}}
\end{figure}

\subsection{The narrow spikes}
I assumed that there is some number $N_{\rm narrow}$ of narrow spikes,
each of which has a gaussian shape. The $i$th narrow spike is parameterised
by its central position $X_{i, \rm narrow}$,
amplitude $A_{i, \rm narrow}$, and width $A_{i, \rm narrow}$.
The parameter vector for narrow spikes is
\begin{align}
\params_{\rm narrow} &=
  \{N_{\rm narrow}, X_1, X_2, ..., X_{N, \rm narrow},
    A_1, A_2, ..., A_{N, \rm narrow},
    W_1, W_2, ..., W_{N, \rm narrow}\}
\end{align}
and the total shape contributed by the
narrow spikes is a sum of the $N_{\rm narrow}$ gaussians:
\begin{align}
f_{\rm narrow}(\x; \params_{\rm narrow}) &=
    \sum_{i=1}^{N_{\rm narrow}} A_{i, \rm narrow}
 \exp\left[-\frac{1}{2W_{i, \rm narrow}^2}\left(x - X_{i, \rm narrow}\right)^2\right].
\end{align}

\subsection{The wide component}
The wide component is also composed of gaussians, like the narrow spikes.
In fact, the only real difference is the prior distributions for the positions,
amplitudes, and widths, given in Section~\ref{sec:priors}.

There is some number $N_{\rm wide}$ of gaussians in the wide component.
The $i$th gaussian is parameterised
by its central position $X_{i, \rm wide}$,
amplitude $A_{i, \rm wide}$, and width $A_{i, \rm wide}$.
The parameter vector for wide gaussians is
\begin{align}
\params_{\rm wide} &=
  \{N_{\rm wide}, X_1, X_2, ..., X_{N, \rm wide},
    A_1, A_2, ..., A_{N, \rm wide},
    W_1, W_2, ..., W_{N, \rm wide}\}
\end{align}
and the total shape contributed by the
wide component is a sum of the $N_{\rm wide}$ gaussians:
\begin{align}
f_{\rm wide}(\x; \params_{\rm wide}) &=
    \sum_{i=1}^{N_{\rm wide}} A_{i, \rm wide}
 \exp\left[-\frac{1}{2W_{i, \rm wide}^2}\left(x - X_{i, \rm wide}\right)^2\right].
\end{align}

Figure~\ref{fig:wide_component} shows several examples of plausible
wide component shapes.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/wide_component.pdf}
\caption{\it Some examples of plausible shapes for the wide component,
generated from the prior distribution for its parameters
(Section~\ref{sec:priors}). For visual purposes only, these
were scaled to amplitudes $\sim$ 1. \label{fig:wide_component}}
\end{figure}

\subsection{Derived properties}
Some derived properties of the wide component are used to summarise its shape.
We have included `skewness', which describes asymmetry, and `nongaussianity'
which quantifies the departure from gaussianity. These are not the same
concept, as the wide component could be non-gaussian while still being symmetric
(and thus having zero skewness).

Defining the normalised version of the wide component as
\begin{align}
g(\x) &= \frac{f_{\rm wide}(\x)}
             {\int_{-\infty}^\infty f_{\rm wide}(\x') \, d\x'},
\end{align}
its center of mass by
\begin{align}
\textnormal{center} &= \int_{-\infty}^\infty x g(\x) \, d\x,
\end{align}
and its width (second central moment) by
\begin{align}
\textnormal{width} &= \sqrt{\int_{-\infty}^\infty
                            g(\x) \left(\x - \textnormal{center}\right)^2
                            \, d\x},
\end{align}
the skewness is
\begin{align}
\textnormal{skewness} &= \int_{-\infty}^\infty
                           g(\x) 
                           \left(
                             \frac{x - \textnormal{center}}{\textnormal{width}}
                           \right)^3
                         \, d\x.
\end{align}

To quantify the non-gaussianity, we construct a gaussian
$g'(\x)$
with the same centre, width, and total integral as $g(\x)$, and compute the
Kullback-Leibler divergence from $g'$ to $g$:
\begin{align}
\textnormal{non-gaussianity} &= 
    \int_{-\infty}^\infty g(\x) \log\left[\frac{g(\x)}{g'(\x)}\right] \, d\x.
\end{align}
where
\begin{align}
g'(\x) &= \frac{1}{\textnormal{width} \times \sqrt{2\pi}}
            \exp\left[-\frac{1}{2\times (\textnormal{width})^2}
                    \left(\x - \textnormal{center}\right)^2\right].
\end{align}
The Kullback-Leibler divergence is a standard way of quantifying how
different one measure is from another \citep{knuth2012foundations}.
It is zero when the two measures are the same, in our case,
if $g(\x)$ is a gaussian.

See Figures~\ref{fig:skewness} and~\ref{fig:nongaussianity} for example
shapes for the wide component, and corresponding values of the skewness
and nongaussianity.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/skewness.pdf}
\caption{\it Examples of wide component shapes with different values
for the skewness.\label{fig:skewness}}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/nongaussianity.pdf}
\caption{\it Examples of wide component shapes with different values
for the nongaussianity.\label{fig:nongaussianity}}
\end{figure}

\section{Prior probability distributions}\label{sec:priors}
The joint prior distribution for the hyperparameters, parameters, and data,
is written $p(\hypers, \params, \data | \info)$, and is usually factorised
in this way:
\begin{align}
p(\hypers, \params, \data | \info) &=
    p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \hypers, \info)\\
    &= p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \info)
\end{align}
where the first step is true in general by the product rule, and the second
step assumes that knowing the parameters would make the hyperparameters
irrelevant when predicting what data would be observed.

All of the assumptions are given in Table~\ref{tab:priors}.

@@

The Laplace, or biexponential, distribution with location parameter
(central position) $a$ and scale parameter (width) $b$ has probability
density given by
\begin{align}
p(x | a, b) &= \frac{1}{2b}\exp\left[-\frac{|x - a|}{b}\right].
\end{align}
We used a Laplace distribution for informative priors, rather than
a more conventional normal distribution, because it assigns higher
prior probability to values far from the center, and can be computed conveniently, i.e., no special functions such as {\tt erf} are needed
to compute the cumulative distribution function.

\section{Installation and usage}

The entire repository, including C++ and Python source code, and a Windows
executable, is hosted on Bitbucket. The URL for the git repository is
For most users, we suggest simply using the pre-compiled Windows executable
{\tt main.exe} from the repository.

The repository can simply be downloaded from

{\tt https://bitbucket.org/eggplantbren/crystals/get/c78decf570f0.zip}

and unzipped into any location on your computer. To execute the program,
run {\tt main.exe}. You will be prompted to enter the filename of the text
file containing the data.

\bibliographystyle{plainnat}
\bibliography{references}

>>>>

\documentclass[review]{elsarticle}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{dsfont}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
% \usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{pdflscape}
\usepackage{gensymb} % For degree symbol
\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\journal{Computers and Geosciences}

\setcitestyle{authoryear,open={(},close={)}}

@@

% http://support.river-valley.com/wiki/index.php?title=Elsarticle.cls

\begin{document}

\begin{frontmatter}

\title{AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction}

\author[rowe]{Michael C. Rowe\corref{cor1}}
\ead{michael.rowe@auckland.ac.nz}
\cortext[cor1]{Corresponding author}
\address[rowe]{School of Environment, The University of Auckland, Private Bag 92019, Auckland 1142, New Zealand}

\author[brewer]{Brendon J. Brewer}
\ead{bj.brewer@auckland.ac.nz}
\address[brewer]{Department of Statistics, The University of Auckland, Private Bag 92019, Auckland 1142, New Zealand}

\begin{abstract}
AMORPH utilizes a new Bayesian statistical approach to interpreting X-ray diffraction results of partially amorphous materials. AMORPH fits X-ray diffraction patterns with a mixture of gaussians,
which allows us to simultaneously infer all of the model parameters and quantify their
uncertainties.
The program simulates background patterns previously manually applied, providing reproducible results, and significantly reducing inter- and intra-user biases. This approach allows for the quantification of amorphous and crystalline materials and for the characterization of the amorphous component, including parameters of centre of mass, width, skewness, and nongaussianity of the amorphous component. Results demonstrate the applicability of this program for calculating amorphous contents of volcanic materials and independently modeling their properties in compositionally variable materials.
\end{abstract}

\begin{keyword}
  Amorphous materials \sep X-ray diffraction \sep Bayesian inference \sep Markov Chain Monte Carlo \sep Mixture models \sep crystallinity
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
Quantification and characterization of amorphous materials is an important area of research in
terrestrial and planetary sciences \citep[e.g.,][]{schmidt2009, dehouck2014, wall2014, morris2016, zorn2017}.
Powder X-ray diffraction (XRD), although primarily a tool for characterization of crystalline materials,
also provides a means to investigate amorphous materials \citep[e.g.,][]{rowe2012, achilles2013}. X-ray diffraction of rocks/soils containing amorphous materials (e.g. volcanic
glass, amorphous silica, and organics) produces a diffraction pattern which is a combination of the
crystalline and amorphous phases. Recent studies have debated the nature of amorphous materials in
planetary sciences \citep[e.g.][]{schmidt2009, dehouck2014}, and a growing number of
applications to terrestrial volcanics have presented an avenue of expansion for XRD methodologies
for amorphous material analysis \citep{ellis2015, kanakiya2017, zorn2017}.
Here we present a new statistical approach and computer program, AMORPH, which utilizes Bayesian inference to determine
the crystallinity, and other characteristics, of partially amorphous materials from X-ray diffraction patterns, as well as the uncertainty in these quantities.

Several techniques currently exist for the quantification of amorphous materials in heterogeneous (amorphous + crystalline) materials. However, these
each present their own limitations. The two most commonly used approaches in recent literature
either rely on a calibration curve for different crystallinities \citep[calibration method;][]{ rowe2012} or a combined Rietveld-Reference Intensity Ratio (RIR) method \citep{gualtieri1996, gualtieri2000}. While the combined Rietveld-RIR procedure allows for quantification of both the amorphous 
and crystalline materials, for proper characterization it requires a library of pure components,
spiked with a reference material (e.g. corundum). Sample spiking presents additional complications
as it reduces the observed intensity of the amorphous material thus reducing the accuracy of
pattern matching to determine abundances. An Excel based adaptation of this pattern matching
technique \citep[FULLPAT;][]{chipera2002} is routinely utilized for characterizing remotely acquired X-ray
diffraction patterns, such as those recovered from the Mars Science Laboratory (MSL) Curiosity rover \citep{bish2013}. The problem faced by RIR approaches is they are based on modeling of an observed 
diffraction pattern compared to library measurements, and therefore do not allow an independent
assessment of the characteristics of the amorphous material. In comparison, the calibration method
focuses solely on the quantification of the amorphous material, utilizing the integrated counts
associated with the amorphous and crystalline componentry.

\begin{align}
C_{\rm m} \% = [C_{\rm pa}]/[A_{\rm pa} + C_{\rm pa}] \times 100
\end{align}

where $C_{\rm m}$ is the measured crystallinity, and $C_{\rm pa}$ and $A_{\rm pa}$ are the integrated peak areas for the
crystalline and amorphous components, respectively (see Figure 5 of \citet{wall2014}). 
While its simplicity has made it advantageous, the calibration methodology has been limited in that
the quality of the results largely depend on the ability of the user to manually assign
backgrounds for count integration, resulting in potentially large inter-user variations. Although the
point of the calibration curve is to reduce the inter-user variation, determinations still tend to
be associated with moderate user error (3-5 vol \%). As every user must independently practice
background fittings using the calibration method, the results ultimately are only as good as the
user’s ability to manually model the data. This technique, while not constrained by a
necessary library of diffraction patterns, also does not characterize the XRD properties of the
amorphous material. This study presents a new approach using Bayesian inference to automate
background determination for the calibration method, removing inter-user and intra-user biases in
the determination of crystallinity and peak characteristics for amorphous/partially amorphous
materials, regardless of what they are.

This paper is structured as follows. In Section~\ref{sec:bayes}, we outline the Bayesian
framework of inference in general terms, and in Section~\ref{sec:model} we describe the
specific model assumptions made by AMORPH (i.e., the mixture of gaussians), and
Section~\ref{sec:priors} lists the probabilistic assumptions (i.e., the prior distributions) of the program. Section~\ref{sec:applications} presents some
example applications and comparisons with standard techniques, and finally we conclude
in Section~\ref{sec:conclusions}. Basic instructions for installation and use of AMORPH are given in~\ref{sec:program}.

\section{Bayesian inference}\label{sec:bayes}
AMORPH is founded upon Bayesian inference
\citep{gregory2005bayesian, o2004kendall, sivia2006data}, where
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Bayesian inference is often implemented computationally with
Markov Chain Monte Carlo (MCMC) methods \citep{mackay2003}, and AMORPH is no exception.
Loosely speaking, the goal is to fit a dataset with a model by exploring
those regions of the model's parameter space that are consistent with the
data. The model assumptions are described in detail in
Section~\ref{sec:model}, and the computation is performed with
Diffusive Nested Sampling, an advanced MCMC method \citep{dns, dnest4}.

Typically, one starts with the `prior distribution'
for unknown quantities (`parameters') $\params$, written $p(\params | \info)$
(the probability distribution for $\params$ {\em given} prior information
$\info$). Throughout this paper, $\params$ is used to denote a
collection of unknown parameters, whereas $2\theta$ (with the factor of two, and the
non-bold $\theta$) is later used to
refer to the incident angle plus reflected angle, which is the
standard notation in X-ray diffraction.

The prior describes an initial state of uncertainty about the
parameter values, and is often a wide probability distribution.
The prior is supplemented with $p(\data | \params, \info)$, sometimes
called a `sampling distribution',
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. This describes uncertainty about what data will be observed,
but encodes some knowledge about the fact that the data has something to do with the
parameters $\params$. For example, $p(\data | \params, \info)$ often encodes
the assumption that the data has noise in it, and the specific noise values
are unknown a priori.

The
product of these two distributions is the joint prior
\begin{align}
p(\params, \data | \info) &= p(\params | \info)p(\data | \params, \info)
\end{align}
which describes uncertainty about the value of both the parameters and the
dataset. Once a specific observed dataset $\data_{\rm obs}$ is known, knowledge about
$\params$ is updated from the prior $p(\params | \info)$ to the posterior
\begin{align}
p(\params | \data_{\rm obs}, \info) &\propto
    p(\params | \info)p(\data_{\rm obs} | \params, \info).
\end{align}
which takes the specific data $\data_{\rm obs}$ into account. With the specific dataset,
$p(\data_{\rm obs} | \params,\info)$ becomes a function of the
unknown parameters $\params$ only and is called the {\em likelihood function}.
The posterior distribution is therefore proportional to the prior distribution
times the likelihood function.

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\phi$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\phi > 3.4$. While many data analysis procedures work by finding the
best fitting values of the parameters (i.e., a maximum likelihood estimate),
Bayesian inference via MCMC involves finding a sampling of plausible values
for the parameters.

\section{The model curve}\label{sec:model}

AMORPH fits the data with a model curve defined over some range from
$x_{\rm min}$ to $x_{\rm max}$ (Fig.~\ref{fig:example_data2}). Throughout this section, we use
$x$ as a synonym for $2\theta$, since the problem is basically a curve-fitting
problem where $x$ and $y$ are typically used.

\begin{figure}[!ht]
\centering
\includegraphics[width=.7\textwidth]{figures/example_data2.jpg}
\caption{\it Powder X-ray diffraction results for a basalt standard with 70\% amorphous material. Annotation corresponds to the model components: background, amorphous (wide) component, and crystalline component (narrow peaks).\label{fig:example_data2}}
\end{figure}

The model curve is a function $y = f(\x; \params)$,
which is assumed to be a sum of the following components identified in Figure~\ref{fig:example_data2}:
\begin{enumerate}
\item A background component, described by parameters $\params_{\rm bg}$
\item An amorphous component, described by parameters $\params_{\rm amorph}$
\item A crystalline component, described by parameters $\params_{\rm xtal}$
\end{enumerate}
The total model function $f_{\rm tot}(x)$ is therefore
\begin{align}
f_{\rm tot}(\x) &= f_{\rm bg}(\x; \params_{\rm bg})
       + f_{\rm amorph}(\x; \params_{\rm amorph})
       + f_{\rm xtal}(\x; \params_{\rm xtal})
\end{align}
and the measured amorphous content ($A_{\rm m}$ or $1-C_{\rm m}$) is defined, similar to Equation 1, as the proportion of the area in the
amorphous and crystalline components which is accounted for by the amorphous component:
\begin{align}
{A_{\rm m} = 1-C_{\rm m}} &= \frac{\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, d\x}
          {\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, dx
            + \int_{\x_{\rm min}}^{\x_{\rm max}} f_{\rm xtal}(\x; \params_{\rm xtal}) \, d\x} \label{eqn:crystallinity}
\end{align}

From each sample from the posterior distribution for all of the parameters,
it is possible to compute the proportion of amorphous material, making it straightforward to
calculate the marginal posterior distribution of the amorphous content, describing
the remaining uncertainty about its value in the light of the data.
In the following subsections, we describe the specific parameterizations of the
three parts of the model curve, that is, the functional forms of
$f_{\rm bg}$, $f_{\rm amorph}$, and $f_{\rm xtal}$.

\subsection{The background component}
The background component is assumed to be piecewise linear, with constant
slope
{between $\x=\x_{\rm min}$ and $\x=10$\degree, between $\x=10$\degree~and $\x=40$\degree,
and between $\x=40$\degree~and $\x=\x_{\rm max}$}. These values are selected to maintain a linear background over the interval from 10--40\degree, the predominant range of amorphous materials analyzed using a Cu K$\alpha$ x-ray source, while still allowing for some flexibility in the structure of the background outside this range.

We parameterized the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which determine the offset of four `control points'
from the mean level. The positions of the four control points are then:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right). \nonumber
\end{align}
The value of the background component at any other point is found by
linear interpolation of these control points.
Figure~\ref{fig:background} shows three example
background curves generated from this parameterization, with typical
values of the parameters.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.6\textwidth]{figures/background.pdf}
\caption{\it Three example functions showing possible shapes of the
background component.\label{fig:background}}
\end{figure}

\subsection{The crystalline component}
We assume that there is some number $N_{\rm xtal}$ of narrow peaks,
each of which has a gaussian shape. The $i$th peak is parameterized
by its central position $X_{i, \rm xtal}$,
amplitude $A_{i, \rm xtal}$, and width $W_{i, \rm xtal}$.
The parameter vector for crystalline component is
\begin{align}
\params_{\rm xtal} &=
  \{N_{\rm xtal}, X_1, X_2, ..., X_{N, \rm xtal},
    A_1, A_2, ..., A_{N, \rm xtal},
    W_1, W_2, ..., W_{N, \rm xtal}\}
\end{align}
and the total shape contributed by the
crystalline component is a sum of the $N_{\rm xtal}$ gaussians:
\begin{align}
f_{\rm xtal}(\x; \params_{\rm xtal}) &=
    \sum_{i=1}^{N_{\rm xtal}} A_{i, \rm xtal}
 \exp\left[-\frac{1}{2W_{i, \rm xtal}^2}\left(x - X_{i, \rm xtal}\right)^2\right].
\end{align}

\subsection{The amorphous component}
The amorphous component is also composed of gaussians, like the crystalline component.
In fact, the only real difference is the prior distributions for the positions,
amplitudes, and widths, given in Section~\ref{sec:priors}.

There is some number $N_{\rm amorph}$ of gaussians in the amorphous component.
The $i$th gaussian is parameterized
by its central position $X_{i, \rm amorph}$,
amplitude $A_{i, \rm amorph}$, and width $W_{i, \rm amorph}$.
The parameter vector for wide gaussians is
\begin{align}
\params_{\rm amorph} &=
  \{N_{\rm amorph}, X_1, X_2, ..., X_{N, \rm amorph},
    A_1, A_2, ..., A_{N, \rm amorph},
    W_1, W_2, ..., W_{N, \rm amorph}\}
\end{align}
and the total shape contributed by the
amorphous component is a sum of the $N_{\rm amorph}$ gaussians:
\begin{align}
f_{\rm amorph}(\x; \params_{\rm amorph}) &=
    \sum_{i=1}^{N_{\rm amorph}} A_{i, \rm amorph}
 \exp\left[-\frac{1}{2W_{i, \rm amorph}^2}\left(x - X_{i, \rm amorph}\right)^2\right].
\end{align}

Figure~\ref{fig:wide_component} shows several examples of plausible amorphous component shapes. The prior distributions (specified in
Section~\ref{sec:priors})
constrain the central positions of these gaussians much
more strongly than the central positions of the narrow peaks (crystalline component), which is how AMORPH is able to distinguish the crystalline
component from the amorphous component.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/wide_component.pdf}
\caption{\it Some examples of plausible shapes for the amorphous component,
generated from the prior distribution for its parameters
(Section~\ref{sec:priors}). For visual purposes only, these
were scaled to amplitudes $\sim$ 1. \label{fig:wide_component}}
\end{figure}

\subsection{Derived properties}
Some derived properties of the amorphous component are used to summarize its shape.
We have included `skewness', which describes asymmetry, and `nongaussianity'
which quantifies the departure from gaussianity. These are not the same
concept, as the amorphous component could be non-gaussian while still being symmetric
(and thus having zero skewness).

Defining the normalized version of the amorphous component as
\begin{align}
g(\x) &= \frac{f_{\rm amorph}(\x)}
             {\int_{-\infty}^\infty f_{\rm amorph}(\x') \, d\x'},
\end{align}
its center of mass by
\begin{align}
\textnormal{center} &= \int_{-\infty}^\infty x g(\x) \, d\x,
\end{align}
and its width (second central moment, perhaps better interpreted
as a half-width) by
\begin{align}
\textnormal{width} &= \sqrt{\int_{-\infty}^\infty
                            g(\x) \left(\x - \textnormal{center}\right)^2
                            \, d\x},
\end{align}
the skewness is
\begin{align}
\textnormal{skewness} &= \int_{-\infty}^\infty
                           g(\x) 
                           \left(
                             \frac{x - \textnormal{center}}{\textnormal{width}}
                           \right)^3
                         \, d\x.
\end{align}

To quantify the non-gaussianity, we construct a gaussian
$g'(\x)$
with the same centre, width, and total integral as $g(\x)$, and compute the
Kullback-Leibler divergence from $g'$ to $g$:
\begin{align}
\textnormal{non-gaussianity} &= 
    \int_{-\infty}^\infty g(\x) \log\left[\frac{g(\x)}{g'(\x)}\right] \, d\x.
\end{align}
where
\begin{align}
g'(\x) &= \frac{1}{\textnormal{width} \times \sqrt{2\pi}}
            \exp\left[-\frac{1}{2\times (\textnormal{width})^2}
                    \left(\x - \textnormal{center}\right)^2\right].
\end{align}
The Kullback-Leibler divergence is a standard way of quantifying how
different one `measure' ($\equiv$ density function, for our purposes here) is from another \citep{knuth2012foundations}.
It is zero when the two measures are the same, in our case,
if $g(\x)$ is a gaussian. See Figures~\ref{fig:skewness} and~\ref{fig:nongaussianity} for example
shapes for the amorphous component, and corresponding values of the skewness
and nongaussianity.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.6\textwidth]{figures/skewness.pdf}
\caption{\it Examples of amorphous component shapes with different values
for the skewness.\label{fig:skewness}}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.6\textwidth]{figures/nongaussianity.pdf}
\caption{\it Examples of amorphous component shapes with different values
for the nongaussianity.\label{fig:nongaussianity}}
\end{figure}

\section{Prior probability distributions}\label{sec:priors}
The joint prior distribution for all of the unknown parameters and
the data must be specified.
When the number of parameters is large, this is often done `hierarchically', by
specifying the priors conditional on some `hyperparameters', and then assigning
priors to the hyperparameters.

The joint prior distribution for
the hyperparameters, parameters, and data,
is written $p(\hypers, \params, \data | \info)$, and is usually factorized
in this way:
\begin{align}
p(\hypers, \params, \data | \info) &=
    p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \hypers, \info)\\
    &= p(\hypers | \info)p(\params | \hypers, \info)
    p(\data | \params, \info)
\end{align}
where the first step is true in general by the product rule, and the second
step assumes that knowing the parameters would make the hyperparameters
irrelevant when predicting what data would be observed. In the AMORPH model,
we introduced hyperparameters describing the typical amplitude
and width of the gaussians for the amorphous component, and
the degree to which the actual amplitudes and widths are scattered
around that typical value. This was also done for the crystalline
component.

A directed acyclic graph, also known as a probabilistic graphical model (PGM),
which depicts the dependence structure of the entire joint
prior probability distribution,
is shown in Figure~\ref{fig:pgm-edited}. The prior distributions themselves
are specified in Table~\ref{tab:priors}.

\begin{figure}[!ht]
\centering
\includegraphics[scale=0.7]{figures/pgm-edited.pdf}
\caption{The PGM. The hyperparameters controlling the prior distributions
for the amorphous and crystalline gaussians are denoted collectively
by $\boldsymbol{\alpha}_{\rm amorph}$ and
$\boldsymbol{\alpha}_{\rm xtal}$ respectively.
The noise-related parameters are collected under $\boldsymbol{\sigma}$.
The data is $\boldsymbol{y}$.\label{fig:pgm-edited}}
\end{figure}

@@

Most of the chosen distributions were intended to be reasonably
conservative and wide.
Some of the prior distributions in AMORPH use the log-uniform distribution,
sometimes (erroneously) called a Jeffreys prior, whose density for a quantity
$z$ is proportional to $1/z$.

We used a Laplace distribution for some informative priors, rather than
a more conventional normal distribution, because it assigns higher
prior probability to values far from the center, and can be computed conveniently, i.e., no special functions such as the inverse of the
error function are needed.
The Laplace, or biexponential, distribution with location parameter
(central position) $a$ and scale parameter (width) $b$ has probability
density given by
\begin{align}
p(x | a, b) &= \frac{1}{2b}\exp\left[-\frac{|x - a|}{b}\right].
\end{align}

The priors for the positions and widths of the `amorphous' and `crystalline'
gaussians are different, to account for the prior knowledge that they
should result in wide gaussians near the centre of the domain, and
narrow gaussians throughout the domain respectively.

The prior for the data given all the parameters is a student-$t$ distribution,
which allows for potentially heavier tails than a normal distribution would.
Thus, analyses with AMORPH should be able to cope with discrepant datapoints
as long as they are not locally correlated (which would cause them to be
mistaken for peaks).

AMORPH does not assume a known number of gaussians for either the
amorphous or the crystalline component. The appropriate number of
gaussians is inferred from the data. The maximum number of gaussians
in each component is set to 300.

\section{Applications}\label{sec:applications}
\subsection{Crystallinity}
The AMORPH program calculates a fraction of amorphous material ($A_{\rm m}$) for each posterior sample of the parameters,
using Equation~\ref{eqn:crystallinity}.Thus measured crystallinity ($C_{\rm m}$) is easily determined from the relationship $C_{\rm m}$ = 1-$A_{\rm m}$, as defined in Equation 5. The measured crystallinity is not equivalent to an "actual" or "true" crystallinity, although the values are often very close to one another.
It should therefore be noted that this procedure does not negate the need for calibration standards
to translate measured crystallinity to an actual crystallinity. Transmission and absorption of X-rays in the material will vary as the ratio of mineral to glass changes in the sample, causing a deviation from a 1:1 line. In addition to theoretical considerations, practical challenges also exist for background fitting. At low crystallinity ($\lesssim 10\%$) diffraction patterns are dominated by the low amplitude amorphous peak such that removing the effects of background noise becomes problematic, resulting in overestimated crystallinities. At high crystallinity ($\gtrsim 90\%$) the problem arises that peak overlaps in multi-mineral assemblages cause broadening effects, and may be subsequently modeled as ``amorphous'' and thus crystallinities  underestimated. The calibration curve therefore is utilized to remove these effects. This is also an important step as
different X-ray diffractometers will provide slightly different calibration curves (Fig.~\ref{fig:intruments}).
Differences between instruments largely reflect changes in background intensity and shape,
particularly at low values of $2\theta$.
